---
title: "PSY4960-Nonparametric Analysis Problem Set"
author: "Amanda Mae Woodward"
date: "11/11/2021"
output: html_document
---
1. Open the diamonds dataset. 
```{r}
data(diamonds)
```

2. What would be the appropriate correlation to calculate to determine the relationship between clarity and price? 
Explain your reasoning. 
  -Cor.test could be good for calculating the correlation along with how significant that correlation is 
  
3. Calculate the correlation you chose in #2. (Hint: you may need to change a class of variable)
```{r}
clarity <- as.numeric(diamonds$clarity)
price <- as.numeric(diamonds$price)
cor.test(clarity, price)

```
4. Calculate a spearman's rho and a kendall's tau for the relationship between carat and clarity. What do you notice about the results?
```{r}
cor.test(clarity, price, method = "spearman")
cor.test(clarity, price, method = "kendall")
#Both have the same P value while the standard deviation is larger for Spearman, and the correlation drawn is also slightly larger with spearman

```
5. Use a Wilcoxon test to compare the price of "preimium" diamonds and "very good" diamonds. How does this compare to the results of a t test?
```{r}
library(tidyr)
cut <- c(diamonds$cut[diamonds$cut == "Very Good"], diamonds$price[diamonds$cut == "Premium"])
levels(cut)  <- c("Very Good", "Premium")

price <- c(diamonds$price[diamonds$cut == "Very Good"], diamonds$price[diamonds$cut == "Premium"])
dat <- cbind.data.frame(cut, price)

t.test(cut, price, data=dat)
wilcox.test(cut, price, data=dat)

#Each one has a similar p value, but the regular t.test offers the a t value, a degrees of freedom, confidence interval, and a mean of x and y while Wilcox offers a W, the smaller of positive and negative ranks
```

6. For the following questions, use the following scenario: 
You are given a 25 question test that you didn't study for and got 15 correct. If you were to completely guess on all questions, would you do better than chance if: 
```{r}
questions <- 25
correct <- 15
#You have a much greater chance at doing well with true/false questions than multiple choice questions of either 3 or 4. 
```
a) it was all true/false questions
```{r}
binom.test(correct, questions, .5)
```
b) it was multiple choice and all questions had 3 choices
```{r}
binom.test(correct, questions, .33)
```
c) it was multiple choice and all questions had 4 choices
```{r}
binom.test(correct, questions, .25)
```
7) open the HairEyeColor dataset
```{r}
data(HairEyeColor)
HairEyeColor<-as.data.frame(HairEyeColor)
# HairEyeColor
```

8) Does red hair occur more often than chance? 
```{r}
hair <- HairEyeColor$Hair
successes <- HairEyeColor$Hair[HairEyeColor$Hair == "Red"]

hairnum <-length(hair) 
successes <- length(successes)

# levels(hair)#Because there are 4 I'm going to say p should be .25



binom.test(successes, hairnum, .25)

#No they do not occur more than chance

```
9) Do Hazel eyes occur more often than chance?

```{r}
eyeslength <- length(HairEyeColor$Eye)
successes <- length(HairEyeColor$Eye[HairEyeColor$Eye == "Hazel"])
# levels(HairEyeColor$Eye)
# eyeslength
binom.test(successes, eyeslength, .25)
#It does not occur more often than chance
```
